---
title: "R Notebook"
output: html_notebook
---
This is a notebook of challenge questions posted into the DiscoRd server

# Reversed Strings

Complete the solution so that it reverses the string passed into it.

Example:
'world'  =>  'dlrow'

Starter Code
solution <- function(s){
  ## your code here
}

```{r}
test <- c("Statistics", "DiscoRd")

reverse.string <- function(x) {
  sapply(lapply(strsplit(x, NULL), rev), paste, collapse = "")
}

reverse.string(test)
```

# 3-7-3/7 
Calculate how many numbers between 1:1000 are divisible by 3, how many are not divisible by 7, and how many are divisible by both 3 and 7.

```{r}
library(tidyverse)

summarise(tibble(v = 1:1000),
  three            = sum(v %% 3 == 0),
  not_seven        = sum(v %% 7 != 0),
  three_and_seven  = sum(v %% 3 == 0 & v %% 7 == 0),
  three_nand_seven = sum(v %% 3 == 0 & v %% 7 != 0)
)
```

# Create an R function to test if a number is a perfect root

```{r}
PerfectRootCheck <- function(input){
  n <- as.integer(input)
  root <- sqrt(n)
  response <- ifelse(as.integer(root + .5)**2 == n, "is a perfect square", "is not a perfect square")
  return(paste(n, response))
}
```

