---
title: "R Notebook"
output: html_notebook
author: "Ryan O'Dea"
---
```{r setup, include = FALSE}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(rstanarm)
library(broom)
```

#Background
Being able to discern between different tastes is very important in the food industry: it enables the development and refinement of new products. Assessing these differences is a common task in the field of sensory discrimination analysis. Suppose that you know well two food products, A and B, to the point of being able to distinguish them by simply trying both. A harder task is to pass a triangle test: you're given two servings of one product and one serving of the other and you're asked to tell which product was different.

PostgreSQL database practice at host mssp1.bu.edu (BU VPN required!) has a table triangle containing some triangle test results for two popular beverages from a number of evaluators. Your task is to conduct a quick statistical analysis of this dataset and summarize your results in a RMarkdown report.


##Warm Up
Connect to the DB as user mssp using

con <- dbConnect(RPostgreSQL::PostgreSQL(), dbname = "practice", host = "mssp1.bu.edu", user = "mssp")

and craft a query to report the proportion of right guesses for each referee.
```{r}
con <- dbConnect(RPostgreSQL::PostgreSQL(), dbname = "practice", host = "mssp1.bu.edu", user = "mssp")

con %>% dbGetQuery("SELECT count(*) FROM triangle") #check

triangle <- tbl(con,"triangle") %>% mutate(y= as.integer(expected == guess))
triangle %>% group_by(referee) %>% summarise(prop=mean(y))

#A better version
con %>% dbGetQuery("SELECT referee, AVG(CAST(expected = guess AS int)) AS prop FROM triangle GROUP BY referee")

```

##Task
Devise a model and fit it to infer if products A and B are distinguishable. In your model, take into account that referees have different taste perception skills.

```{r}
triangle <- con %>% dbGetQuery("SELECT * FROM triangle")
dbDisconnect(con)

triangle <- triangle %>% mutate(y = as.integer(expected == guess), referee=factor(referee))

contrasts(triangle$referee) <-contr.sum
glm(y~referee,family=binomial, data=triangle)

```


##Extras

1. Investigate via a model if there's a difference in test outcomes if either product is prevalent (A or B).

2. Investigate via a model if there's a difference in test outcomes if sequence order matters (e.g., AAB vs ABA vs BAA).
