---
title: "Decision Trees"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=7 ,out.width="0.7\\linewidth",fig.align = "center")
pacman::p_load(C50,ggplot2,tree,BayesTree,RColorBrewer,randomForest,rpart.plot,pROC)
```

# Fitting Classification Trees
The tree library is used to construct classification and regression trees. 

```{r}
pacman::p_load(caret,gbm,tree,ISLR,MASS,party,partykit,pROC,randomForest,rpart,rpart.plot)
```

We first use classification trees to analyze the `Carseats` data set. In these data, Sales is a continuous variable, and so we begin by recoding it as a binary variable. We use the `ifelse()` function to create a variable, called `High`, which takes on a value of `Yes` if the `Sales` variable exceeds `8`, and takes on a value of `No` otherwise.

```{r}
library(ISLR)
attach(Carseats)
High=as.factor(ifelse(Sales<=8,"No","Yes"))
```

Finally, we use the data.frame() function to merge High with the rest of the Carseats data.
```{r}
Carseats=data.frame(Carseats,High)
```

We now use the `tree()` function to fit a classification tree in order to predict High using all variables but Sales. The syntax of the `tree()` function is quite similar to that of the `lm()` function.
```{r}
tree.carseats=tree(High~.-Sales,Carseats)
```

The `summary()` function lists the variables that are used as internal nodes
in the tree, the number of terminal nodes, and the (training) error rate.

```{r}
summary(tree.carseats)
```


We see that the training error rate is $9\%$. For classification trees, the deviance reported in the output of `summary()` is given by

$$-2\sum_m \sum_k n_{mk}\log\hat{p}_{mk},$$
where $n_{mk}$ is the number of observations in the mth terminal node that belong to the $k$th class. A small deviance indicates a tree that provides a good fit to the (training) data. The residual mean deviance reported is simply the deviance divided by $n-|T_0|$, which in this case is $400-27 = 373$.

One of the most attractive properties of trees is that they can be graphically displayed. We use the `plot()` function to display the tree structure, and the `text()` function to display the node labels. The argument `pretty=0` instructs R to include the category names for any qualitative predictors, rather than simply displaying a letter for each category.

```{r,fig.width=12,fig.height=10}
plot(tree.carseats)
text(tree.carseats,pretty=0)
```

The most important indicator of `Sales` appears to be shelving location, since the first branch differentiates Good locations from Bad and Medium locations.

If we just type the name of the tree object, R prints output corresponding to each branch of the tree. R displays the split criterion (e.g. `Price<92.5`), the number of observations in that branch, the deviance, the overall prediction for the branch (`Yes` or `No`), and the fraction of observations in that branch that take on values of `Yes` and `No`. Branches that lead to terminal nodes are indicated using asterisks.

```{r}
tree.carseats
```

In order to properly evaluate the performance of a classification tree on these data, we must estimate the test error rather than simply computing the training error. We split the observations into a training set and a test set, build the tree using the training set, and evaluate its performance on the test data. The `predict()` function can be used for this purpose. In the case of a classification tree, the argument `type="class"` instructs R to return the actual class prediction. This approach leads to correct predictions for around 71.5 % of the locations in the test data set.

```{r}
set.seed(2021);train=sample(1:nrow(Carseats), 200)
Carseats.test=Carseats[-train,]
High.test=High[-train]
tree.carseats=tree(High~.-Sales,Carseats,subset=train)
tree.pred=predict(tree.carseats,Carseats.test,type="class")
table(tree.pred,High.test)
(99+51)/200
```

Next, we consider whether pruning the tree might lead to improved results. The function `cv.tree()` performs cross-validation in order to determine the optimal level of tree complexity; cost complexity pruning is used in order to select a sequence of trees for consideration. We use the argument FUN=prune.misclass in order to indicate that we want the classification error rate to guide the cross-validation and pruning process, rather than the default for the `cv.tree()` function, which is deviance. The `cv.tree()` function reports the number of terminal nodes of each tree con- sidered (size) as well as the corresponding error rate and the value of the cost-complexity parameter used (k, which corresponds to $\alpha$ in (8.4)).

```{r}
set.seed(2021);cv.carseats=cv.tree(tree.carseats,FUN=prune.misclass)
names(cv.carseats)
cv.carseats
```
Note that, despite the name, dev corresponds to the cross-validation error rate in this instance. The tree with 9 terminal nodes results in the lowest cross-validation error rate, with 50 cross-validation errors. We plot the error rate as a function of both size and k.

```{r,fig.width=7,fig.height=3}
par(mfrow=c(1,2))
plot(cv.carseats$size,cv.carseats$dev,type="b")
plot(cv.carseats$k,cv.carseats$dev,type="b")
```

We now apply the `prune.misclass()` function in order to prune the tree to obtain the nine-node tree.
```{r}
prune.carseats=prune.misclass(tree.carseats,best=9)
plot(prune.carseats);text(prune.carseats,pretty=0)
```

How well does this pruned tree perform on the test data set? Once again, we apply the `predict()` function.
```{r}
tree.pred=predict(prune.carseats,Carseats.test,type="class")
table(tree.pred,High.test)
(101+47)/200
```

Now 77 % of the test observations are correctly classified, so not only has the pruning process produced a more interpretable tree, but it has also improved the classification accuracy.
If we increase the value of best, we obtain a larger pruned tree with lower classification accuracy:

```{r}
prune.carseats=prune.misclass(tree.carseats,best=15)
plot(prune.carseats);text(prune.carseats,pretty=0)
tree.pred=predict(prune.carseats,Carseats.test,type="class")
table(tree.pred,High.test)
(99+51)/200
```

# Fitting Regression Trees

Here we fit a regression tree to the Boston data set. First, we create a training set, and fit the tree to the training data.

```{r}
library(MASS)
set.seed(1);train = sample(1:nrow(Boston), nrow(Boston)/2)
tree.boston=tree(medv~.,Boston,subset=train)
summary(tree.boston)
```
Notice that the output of `summary()` indicates that only three of the variables have been used in constructing the tree. In the context of a regression tree, the deviance is simply the sum of squared errors for the tree. We now plot the tree.

```{r}
plot(tree.boston);text(tree.boston,pretty=0)
```

The variable `lstat` measures the percentage of individuals with lower socioeconomic status. The tree indicates that lower values of lstat correspond to more expensive houses. The tree predicts a median house price of $46, 400 for larger homes in suburbs in which residents have high socioeconomic status (`rm>=7.437` and `lstat<9.715`).
Now we use the `cv.tree()` function to see whether pruning the tree will improve performance.

```{r}
cv.boston=cv.tree(tree.boston)
plot(cv.boston$size,cv.boston$dev,type='b')
```

In this case, the most complex tree is selected by cross-validation. However, if we wish to prune the tree, we could do so as follows, using the `prune.tree()` function:

```{r}
prune.boston=prune.tree(tree.boston,best=7)
plot(prune.boston);text(prune.boston,pretty=0)
```

In keeping with the cross-validation results, we use the unpruned tree to make predictions on the test set.

```{r}
yhat=predict(tree.boston,newdata=Boston[-train,])
boston.test=Boston[-train,"medv"]
plot(yhat,boston.test);abline(0,1)
sqrt(mean((yhat-boston.test)^2))
```

In other words, the test set MSE associated with the regression tree is 25.05. The square root of the MSE is therefore around 5.005, indicating that this model leads to test predictions that are within around $5, 005 of the true median home value for the suburb.

## rpart plots

package rpart provide nice plots for trees.  Here are examples using the iris data.

```{r}

rpart <- rpart(Species ~ ., data=iris, method="class",)
rpart.plot(rpart)
prp(rpart)
```

# Bagging and Random Forests

Here we apply bagging and random forests to the Boston data, using the randomForest package in R. The exact results obtained in this section may depend on the version of R and the version of the `randomForest` package installed on your computer. Recall that bagging is simply a special case of a random forest with `m = p`. Therefore, the `randomForest()` function can be used to perform both random forests and bagging. We perform bagging as follows:

```{r}

set.seed(1)
bag.boston=randomForest(medv~.,data=Boston,
                        subset=train,
                        mtry=13,
                        importance=TRUE)
bag.boston
```
The argument `mtry=13` indicates that all 13 predictors should be considered for each split of the tree—-in other words, that bagging should be done. How well does this bagged model perform on the test set?


```{r}
yhat.bag = predict(bag.boston,newdata=Boston[-train,])
plot(yhat.bag, boston.test);abline(0,1)
mean((yhat.bag-boston.test)^2)
```

The test set MSE associated with the bagged regression tree is `r round(mean((yhat.bag-boston.test)^2),2)`, almost half that obtained using an optimally-pruned single tree. We could change the number of trees grown by `randomForest()` using the ntree argument:

```{r}
bag.boston=randomForest(medv~.,
                        data=Boston,
                        subset=train,
                        mtry=13,
                        ntree=25)
yhat.bag = predict(bag.boston,newdata=Boston[-train,])
mean((yhat.bag-boston.test)^2)
bag.boston
```

Growing a random forest proceeds in exactly the same way, except that we use a smaller value of the mtry argument. By default, `randomForest()` uses $p/3$ variables when building a random forest of regression trees, and $\sqrt{p}$ variables when building a random forest of classification trees. Here we use `mtry = 6`.  

```{r}
set.seed(1)
rf.boston=randomForest(medv~.,
                       data       = Boston,
                       subset     = train,
                       mtry       = 6,
                       importance = TRUE)
yhat.rf = predict(rf.boston, newdata=Boston[-train,])
mean((yhat.rf-boston.test)^2)
```

The test set MSE is `r round(mean((yhat.rf-boston.test)^2),2)`; this indicates that random forests yielded an improvement over bagging in this case.
Using the `importance()` function, we can view the importance of each variable.

```{r}
importance(rf.boston)
```

Two measures of variable importance are reported. The former is based upon the mean decrease of accuracy in predictions on the out of bag samples when a given variable is excluded from the model. The latter is a measure of the total decrease in node impurity that results from splits over that variable, averaged over all trees (this was plotted in Figure 8.9). In the case of regression trees, the node impurity is measured by the training RSS, and for classification trees by the deviance. Plots of these importance measures can be produced using the `varImpPlot()` function.

```{r,fig.width=8, fig.height=4}
varImpPlot(rf.boston)
```

The results indicate that across all of the trees considered in the random forest, the wealth level of the community (`lstat`) and the house size (`rm`) are by far the two most important variables.

# Boosting

Here we use the `gbm` package, and within it the `gbm()` function, to fit boosted regression trees to the Boston data set. We run `gbm()` with the option `distribution="gaussian"` since this is a regression problem; if it were a binary classification problem, we would use `distribution="bernoulli"`. The argument `n.trees=5000` indicates that we want 5000 trees, and the option `interaction.depth=4` limits the depth of each tree.

```{r}

set.seed(1)
boost.boston = gbm(medv~.,
                 data=Boston[train,],
                 distribution="gaussian",
                 n.trees=5000,
                 interaction.depth=4)
```

The `summary()` function produces a relative influence plot and also outputs the relative influence statistics.

```{r,fig.width=6,fig.height=5}
summary(boost.boston,plotit=FALSE)
plot.gbm.tree<-function(object){
  summary.object<-summary(object,plotit=FALSE)
  reorder_size <- function(x) {
    factor(x, levels = names(sort(table(x))))
  }
  dttmp<-summary.object
  dttmp$variables <- factor(dttmp$var, 
                            levels = dttmp$var[order(dttmp$rel.inf)])
  colourCount = length(dttmp$variables)
  color_levels<-colorRampPalette(brewer.pal(9, "Blues"))(colourCount)
  gp<- ggplot(dttmp) + geom_bar() + 
  aes( x=variables, weight = rel.inf , fill=variables)+
  ylab("Relative influence")+ coord_flip()+
   scale_fill_manual(values = color_levels )+ theme(legend.position="none")
  print(gp)
}
plot.gbm.tree(boost.boston)
```

We see that `lstat` and `rm` are by far the most important variables. We can also produce partial dependence plots for these two variables. These plots illustrate the marginal effect of the selected variables on the response after integrating out the other variables. In this case, as we might expect, median house prices are increasing with rm and decreasing with lstat.

```{r,fig.width=8,fig.height=4}
par(mfrow=c(1,2))
plot(boost.boston,i="rm");plot(boost.boston,i="lstat")
```

We now use the boosted model to predict medv on the test set:

```{r}
yhat.boost=predict(boost.boston,
                   newdata=Boston[-train,],
                   n.trees=5000)
mean((yhat.boost-boston.test)^2)
```

The test MSE obtained is 11.8; similar to the test MSE for random forests and superior to that for bagging. If we want to, we can perform boosting with a different value of the shrinkage parameter $\lambda$ in (8.10). The default value is 0.001, but this is easily modified. Here we take $\lambda = 0.2$.

```{r}
boost.boston=gbm(medv~.,
                 data=Boston[train,],
                 distribution="gaussian",
                 n.trees=5000,
                 interaction.depth=4,
                 shrinkage=0.2,
                 verbose=F)
yhat.boost=predict(boost.boston,newdata=Boston[-train,],n.trees=5000)
mean((yhat.boost-boston.test)^2)
```

In this case, using $\lambda = 0.2$ leads to a slightly lower test MSE than $\lambda = 0.001$.

# Extra

https://cran.r-project.org/web/packages/partykit/vignettes/ctree.pdf

```{r,fig.width=12,fig.height=10}
# Script to compare ctree with rpart

# Cell Body Segmentation

data <- iris

trainIndex <- createDataPartition(data$Species,p=.7,list=FALSE)
trainData <- data[trainIndex,]
testData  <- data[-trainIndex,]
#------------------------
set.seed(23)
# Fit Conditional Tree Model
ctree.fit <- partykit::ctree(Species ~ ., data=trainData)
ctree.fit	
plot(ctree.fit,main="ctree Model")
```


```{r,fig.width=8,fig.height=8}
#Make predictions using the test data set
ctree.pred <- predict(ctree.fit,testData)
#Draw the ROC curve 
ctree.ROC <- roc(predictor=as.numeric(ctree.pred),
                 response=testData$Species)
ctree.ROC$auc
#Area under the curve: 0.8326
plot(ctree.ROC,main="ctree ROC")
```

```{r,fig.width=12,fig.height=10}
# Fit CART  Model
rpart.fit <- rpart(Species ~ ., data=trainData,cp=0)
rpart.fit		
plot(as.party(rpart.fit),main="rpart Model")
```

```{r}
#Make predictions using the test data set
rpart.pred <- predict(rpart.fit,newdata=testData,type="class")
#Draw the ROC curve 
rpart.ROC <- roc(predictor=as.numeric(rpart.pred),
                 response=testData$Species)
rpart.ROC$auc
#Area under the curve: 0.8536
plot(rpart.ROC)
```

## C4.5 

C4.5 is similar to CART and it was considered the best around 2008.
C4.5 is different from CART in the following ways
- C4.5 uses Shannon Entropy to pick features with the greatest information gain as nodes.
- CART uses Gini Impurity
Problem is I was not able to get RWeka to work on OS X.
So I will skip this section and look at C5.0 which is improvement on C4.5.

```{r, eval=FALSE}
# load the package
library(RWeka)
# fit model
fit4.5 <- J48(Species~., data=iris)
# summarize the fit
summary(fit4.5)
# make predictions
predictions <- predict(fit4.5, iris[,1:4])
# summarize accuracy
table(predictions, fit4.5$Species)

## PART

# fit model
fit <- PART(Species~., data=iris)
# summarize the fit
summary(fit)
# make predictions
predictions <- predict(fit, iris[,1:4])
# summarize accuracy
table(predictions, iris$Species)
```

## C5.0

C5.0 is an extention of C4.5.  It is said to be
- faster than C4.5 (several orders of magnitude)
- more memory efficient than C4.5
- Smaller decision trees compared to C4.5
- Can perform boosting to improve accuracy.
- Weighting: allows you to weight different cases and misclassification types.
- Winnowing: automatically winnows the attributes to remove those that may be unhelpful.

```{r}
#http://machinelearningmastery.com/non-linear-classification-in-r-with-decision-trees/
# load the package
library(C50)
# fit model
fit1 <- C5.0(Species~., data=iris)
# fit boosted model
fitb <- C5.0(Species~., data=iris, trials=10)
# summarize the fit
print(fit1)
# make predictions
predictions1 <- predict(fit1, iris)
```

### summarize accuracy 
With boosting
```{r}
table(predictions1, iris$Species)
```

With boosting
```{r}
table(predict(fitb, iris), iris$Species)
```


# BART
BART is a Bayesian “sum-of-trees” model.
For numeric response $y$, we have $y_i = f(x_i) + e_i$, where $e_i ~ N(0,\sigma^2)$.
For a binary response $y$, $P(Y=1 | x) = F(f(x))$, 
where $F$ denotes the standard normal cdf (probit link).

In both cases, $f$ is the sum of many tree models. 
The goal is to have very flexible inference for the uknown function $f$.

In the spirit of “ensemble models”, each tree is constrained by a prior to be a weak learner so that it contributes a small amount to the overall fit.
```{r}
library(BayesTree)
iris_sub<-iris[iris$Species %in% c("setosa" ,"versicolor"),]
iris_sub$Species<-droplevels(iris_sub$Species)
iris_test_idx<- sample(1:nrow(iris_sub),80)
iris_test_spec_01<-as.integer(iris_sub$Species[-iris_test_idx])-1
bartFit = bart( x.train=iris_sub[iris_test_idx,-5],
                y.train=iris_sub$Species[iris_test_idx],
                x.test=iris_sub[-iris_test_idx,-5],
                ndpost=200) 
mean(1*(pnorm(apply(bartFit$yhat.test,2,mean))>0.5)!=iris_test_spec_01 )
plot(bartFit) # plot bart fit
# library(bartMachine)
```